---
title: "import and wrangle data"
format: html
editor: visual
---

# Setup and download data from DNAnexus

Before running this, wipe your environment clean with the broom to the right. Find the name of your exported file on DNAnexus in your project.

```{r}
install.packages("here", "dplyr", "tidyr", "purrr", "stringr", "lubridate", "janitor")

library(dplyr)
library(tidyr)
library(purrr)
library(lubridate)
library(stringr)

system("dx download table-exporter_2025-03-18_09-45-45_data.csv", intern = TRUE) 

dataset <- read.csv(here::here("data-raw", "table-exporter_2025-03-18_09-45-45_data.csv"))
```

# Clean column names

```{r}
dataset <- janitor::clean_names(dataset)

```

# Look at the data structure

```{r}
dataset %>% select(starts_with("diagnoses_icd10"), starts_with("date_of_first_in_patient_diagnosis_icd10")) %>% head()
```

# Function to split icd and operative procedure columns

```{r}
split_columns <- function(data, icd_col_name) {
    max_instances <- data[[icd_col_name]] %>%
        replace_na("") %>%
        str_count("\\|") %>%  # Count the number of pipes
        max(na.rm = TRUE) + 1  # Get the max number of splits plus 1

    # Perform the separation based on the pipe operator
    data <- data %>%
        separate(
            icd_col_name,
            into = paste0(icd_col_name, "_array_", 0:(max_instances - 1)),
            sep = "\\|",
            fill = "right"
        )

    return(data)
}

```

# Apply the function to both ICD10 and ICD9 columns

```{r}
dataset <- dataset %>%
    split_columns("diagnoses_icd10") %>%
    split_columns("diagnoses_icd9") %>%
    split_columns("operative_procedures_opcs4") %>%
    split_columns("operative_procedures_opcs3")
```

Look at the data structure again

# Specify your ICD10 codes:

```{r}
icd10_codes <- c("DE10.5", "DE11.5", "DE13.5", "DE14.5", "DG45.0", "DG45.1", "DG45.3", "DG45.4", "DG45.8", "DG45.9", "DI20", "DI21",
                 "DI22.0", "DI22.1", "DI22.9", "DI23", "DI24.0", "DI24.1", "DI24.8", "DI24.9", "DI25", "DI60", "DI64.9", "DI64.9", "DI69.3",
                 "DI69.4", "DI69.8", "DI70.2", "DI70.2A", "DI73.9", "DI979B", "DI979E", "E10.5", "E11.5", "E13.5", "E14.5", "I20.0",
                 "I20.9", "I21.0", "I21.1", "I21.2", "I21.3", "I21.4", "I21.9", "I24.8", "I24.9", "I25.0", "I25.1", "I25.5", "I25.8",
                 "I25.9", "I60.6", "I60.9", "I63.4", "I63.6", "I63.8", "I63.9", "I64", "I70.2")

icd10_codes <-  str_replace(icd10_codes, "^D", "")


# regex pattern to match in the icd10 to icd9 table below
pattern <- paste0("^(", paste(icd10_codes, collapse = "|"), ")")


```

# Translate into ICD9 codes

Using a table found on <https://www.nber.org/research/data/icd-9-cm-and-icd-10-cm-and-icd-10-pcs-crosswalk-or-general-equivalence-mappings>

```{r}
icd10to9 <- read.csv("https://data.nber.org/gem/icd10cmtoicd9gem.csv") %>%
 mutate(across(everything(), ~ str_replace_all(., '"', ''))) %>%
  select(icd10 = icd10cm, icd9 = icd9cm) %>%
  mutate(icd10 = str_replace(icd10, "(\\d{2})(\\D.*)", "\\1.\\2"),
         icd10 = str_replace(icd10, "(\\d{2})(\\d.*)", "\\1.\\2"))

icd9_codes <- icd10to9 %>%
  filter(str_detect(icd10, pattern)) %>%
  pull(icd9) %>% 
    unique()
```

# Specify operative procedure codes

```{r}
opcs4_codes <- c("K40", "K41", "K42", "K43", "K44", "K45", "K46", "K48", "K49", "K50", "K75")
```

# Function to process ICD and OPCS codes

```{r}
process_codes <- function(
    codes = icd10_codes, 
    prefix = "diagnoses_icd10", 
    date_prefix = "date_of_first_in_patient_diagnosis_icd10", 
    row_range, 
    new_col_name = "ascvd"  # New argument for dynamic column name
) {
  dataset[row_range, ] %>%
    select(participant_id, starts_with(prefix), starts_with(date_prefix)) %>%
    pivot_longer(
      cols = -participant_id,
      names_to = c(".value", "array_index"),
      names_pattern = "(.*)_array_(\\d+)"
    ) %>%
    filter(!is.na(!!sym(prefix)), !!sym(prefix) != "") %>%
    mutate(
      !!sym(new_col_name) := case_when(
        map_lgl(!!sym(prefix), ~ any(str_starts(.x, codes))) ~ 1,  # Check if any code matches
        TRUE ~ 0
      )
    ) %>%
    filter(!!sym(new_col_name) == 1) %>%
    rename(
      dia_proc = !!sym(prefix),  # Rename diagnoses_icd10 to diagnoses_procedures
      date = !!sym(paste0(date_prefix))  # Rename date columns to 'date'
    )
}
```

# Batch process

Depending on the number of diagnosis codes and the number of individuals with a matching code, it might be necessary to split the batches even further.

```{r}

icd10_batch1 <- process_codes(codes = icd10_codes, prefix = "diagnoses_icd10", row_range = 1:200000)

icd10_batch2 <- process_codes(codes = icd10_codes, prefix = "diagnoses_icd10", row_range = 200001:400000)

icd10_batch3 <- process_codes(codes = icd10_codes, prefix = "diagnoses_icd10", row_range = 400001:502129)

icd9_batch1 <- process_codes(codes = icd9_codes, prefix = "diagnoses_icd9", row_range = 1:502129) 

opcs4_batch1 <- process_codes(codes = opcs4_codes, prefix = "operative_procedures_opcs4", row_range = 1:502129 )

```

# Create diagnosis

Choose either to run all above or test out with only one.

```{r}
# ascvd_col <- bind_rows(icd10_batch1, icd10_batch2, icd10_batch3, icd9_batch1, opcs4_batch1) %>%

ascvd_col <- icd10_batch1 %>% 
  group_by(participant_id) %>%
    arrange(date) %>%  # Ensure data is sorted by date for each participant
    slice_head(n = 1) %>%  # Keep only the first row per participant (earliest date)
    ungroup() %>%
    mutate(date = as_date(date)) %>%
    rename(
        ascvd_date = date,
        ascvd_array_index = array_index,
        ascvd_source = dia_proc
    ) %>%
    arrange(participant_id)  # Final sorting by participant_id

```

# Join ascvd_col to the main data and fill in the missing information

```{r}

ascvd_col <- ascvd_col %>% 
  mutate(date = as_date(date)) %>% 
  rename(ascvd_date =date,
         ascvd_array_index = array_index,
         ascvd_dia_proc = dia_proc)
```

Censoring date set to 2022-05-31:

```{r}

dataset_ascvd <- dataset %>% 
  select(-starts_with("diagnoses"), -starts_with("date_of_first"), -starts_with("operative_procedures")) %>%
  left_join(ascvd_col) %>% 
  mutate(date_lost_to_follow_up = as_date(date_lost_to_follow_up),
    ascvd = case_when(
    is.na(ascvd) ~ 0,
    ascvd_date > as_date("2022-05-31") ~ 0,
    ascvd == 1 & as_date(date_lost_to_follow_up)<ascvd_date ~ 0,
    TRUE ~ ascvd),
    ascvd_date = case_when(
      ascvd==0 ~ as_date("2022-05-31"),
      is.na(ascvd_date) & ascvd == 1 ~ NA,
      TRUE ~ ascvd_date
    ),
    ascvd_date = case_when(
      date_lost_to_follow_up < ascvd_date ~ date_lost_to_follow_up,
      ascvd_date > as_date("2022-05-31") ~ as_date("2022-05-31"),
      TRUE ~ ascvd_date
    )
  )
```

#Upload to DNAnexus

```{r}
write.csv(dataset_ascvd, "dataset_ascvd.csv")

system("dx upload dataset_ascvd.csv") 

```
