---
title: "import and wrangle data"
format: html
editor: visual
---

# Setup and download data from DNAnexus

Before running this, wipe your environment clean with the broom to the right. Find the name of your exported file on DNAnexus in your project.

```{r}
install.packages(c("here", "dplyr", "tidyr", "purrr", "stringr", "lubridate", "janitor"))

library(dplyr)
library(tidyr)
library(purrr)
library(lubridate)
library(stringr)

system("dx download table-exporter_2025-03-18_09-45-45_data.csv", intern = TRUE) 

dataset <- read.csv(here::here("data-raw", "table-exporter_2025-03-18_09-45-45_data.csv"))
```

# Clean column names

```{r}
dataset <- janitor::clean_names(dataset)

```

# Look at the data structure

```{r}
dataset %>% select(starts_with("diagnoses_icd10"), starts_with("date_of_first_in_patient_diagnosis_icd10")) %>% head()
```

# Function to split icd and operative procedure columns

The function below is based on code provided by a UKB data analyst:\
<https://community.ukbiobank.ac.uk/hc/en-gb/community/posts/23567724254621-If-I-want-to-obtain-the-first-diagnosis-date-of-patients-with-ICD-code-K70-0-Alcoholic-fatty-liver-from-the-Summary-Diagnoses-which-online-tool-should-I-use>

```{r}
split_columns <- function(data, icd_col_name) {
    max_instances <- data[[icd_col_name]] %>%
        replace_na("") %>%
        str_count("\\|") %>%  # Count the number of pipes
        max(na.rm = TRUE) + 1  # Get the max number of splits plus 1

    # Perform the separation based on the pipe operator
    data <- data %>%
        separate(
            icd_col_name,
            into = paste0(icd_col_name, "_array_", 0:(max_instances - 1)),
            sep = "\\|",
            fill = "right"
        )

    return(data)
}

```

# Apply the function to both ICD10 and ICD9 columns

```{r}
dataset <- dataset %>%
    split_columns("diagnoses_icd10") %>%
    split_columns("diagnoses_icd9") %>%
    split_columns("operative_procedures_opcs4") %>%
    split_columns("operative_procedures_opcs3")
```

Look at the data structure again:

```{r}
dataset %>% select(starts_with("diagnoses_icd10"), starts_with("date_of_first_in_patient_diagnosis_icd10")) %>% head()
```

You have now successfully split you ICD and operative procedure columns and can now save the split dataset to DNAnexus for further processing in your desired program:

```{r}

# write.csv(dataset, "dataset_split.csv")
#  
# system("dx upload dataset_split.csv") 

```

Or you can continue to play with creating your endpoint:

# Specify your ICD10 codes

DISCLAIMER: This next section is entirely based on the code for my personal project. It is not meant to be a recipe for how to create an endpoint but only an example of a way to do it. Please always consider your own course of action in regard to selecting relevant ICD and procedure codes, dates and censoring criteria.

In this example we will create an ascvd endpoint based on the ICD10 codes that I used to makes a similar endpoint in CGPS data:

```{r}
icd10_codes <- c("DE10.5", "DE11.5", "DE13.5", "DE14.5", "DG45.0", "DG45.1", "DG45.3", "DG45.4", "DG45.8", "DG45.9", "DI20", "DI21",
                 "DI22.0", "DI22.1", "DI22.9", "DI23", "DI24.0", "DI24.1", "DI24.8", "DI24.9", "DI25", "DI60", "DI64.9", "DI64.9", "DI69.3",
                 "DI69.4", "DI69.8", "DI70.2", "DI70.2A", "DI73.9", "DI979B", "DI979E", "E10.5", "E11.5", "E13.5", "E14.5", "I20.0",
                 "I20.9", "I21.0", "I21.1", "I21.2", "I21.3", "I21.4", "I21.9", "I24.8", "I24.9", "I25.0", "I25.1", "I25.5", "I25.8",
                 "I25.9", "I60.6", "I60.9", "I63.4", "I63.6", "I63.8", "I63.9", "I64", "I70.2")

#remove the D
icd10_codes <-  str_replace(icd10_codes, "^D", "")


# regex pattern to match in the icd10 to icd9 table below
pattern <- paste0("^(", paste(icd10_codes, collapse = "|"), ")")


```

# Translate into ICD9 codes

Using a table found on <https://www.nber.org/research/data/icd-9-cm-and-icd-10-cm-and-icd-10-pcs-crosswalk-or-general-equivalence-mappings>.

The UK Biobank also provides a conversion table: <https://biobank.ndph.ox.ac.uk/ukb/coding.cgi?id=1836>

This is however only a partial mapping, that does not take into account subdiagnoses of ICD10 codes.

```{r}
icd10to9 <- read.csv("https://data.nber.org/gem/icd10cmtoicd9gem.csv") %>%
  select(icd10 = icd10cm, icd9 = icd9cm) %>% #only select icd10 and 9 columns
  mutate(icd10 = str_replace(icd10, "(\\d{2})(\\d.*)", "\\1.\\2")) #insert a . after the first two numeric characters to match the style of our data

icd9_codes <- icd10to9 %>%
  filter(str_detect(icd10, pattern)) %>%
  pull(icd9) %>% 
    unique()
```

# Specify operative procedure codes

```{r}
opcs4_codes <- c("K40", "K41", "K42", "K43", "K44", "K45", "K46", "K48", "K49", "K50", "K75")
```

This can of course also be done for opcs3 codes. This is however not relevant for the ascvd diagnosis we are creating, as there are no events.

# Function to process ICD and OPCS codes

```{r}
process_codes <- function(
    codes = icd10_codes, 
    prefix = "diagnoses_icd10", 
    date_prefix = "date_of_first_in_patient_diagnosis_icd10", 
    row_range, 
    new_col_name = "ascvd"  # New argument for dynamic column name
) {
  dataset[row_range, ] %>%
    select(participant_id, starts_with(prefix), starts_with(date_prefix)) %>%
    pivot_longer(
      cols = -participant_id,
      names_to = c(".value", "array_index"),
      names_pattern = "(.*)_array_(\\d+)"
    ) %>%
    filter(!is.na(!!sym(prefix)), !!sym(prefix) != "") %>%
    mutate(
      !!sym(new_col_name) := case_when(
        map_lgl(!!sym(prefix), ~ any(str_starts(.x, codes))) ~ 1,  # Check if any code matches
        TRUE ~ 0
      )
    ) %>%
    filter(!!sym(new_col_name) == 1) %>%
    rename(
      dia_proc = !!sym(prefix),  # Rename diagnoses_icd10 to diagnoses_procedures
      date = !!sym(paste0(date_prefix))  # Rename date columns to 'date'
    )
}
```

# Batch process

Depending on the number of diagnosis codes and the number of individuals with a matching code, it might be necessary to split the batches even further.

```{r}

icd10_batch1 <- process_codes(codes = icd10_codes, prefix = "diagnoses_icd10", row_range = 1:200000)

icd10_batch2 <- process_codes(codes = icd10_codes, prefix = "diagnoses_icd10", row_range = 200001:400000)

icd10_batch3 <- process_codes(codes = icd10_codes, prefix = "diagnoses_icd10", row_range = 400001:502129)

icd9_batch1 <- process_codes(codes = icd9_codes, prefix = "diagnoses_icd9", row_range = 1:502129) 

opcs4_batch1 <- process_codes(codes = opcs4_codes, prefix = "operative_procedures_opcs4", row_range = 1:502129 )

```

# Create diagnosis

Choose either to run all above or test out with only one.

```{r}
# ascvd_col <- bind_rows(icd10_batch1, icd10_batch2, icd10_batch3, icd9_batch1, opcs4_batch1) %>%

ascvd_col <- icd10_batch1 %>% 
  group_by(participant_id) %>%
    arrange(date) %>%  # Ensure data is sorted by date for each participant
    slice_head(n = 1) %>%  # Keep only the first row per participant (earliest date)
    ungroup() %>%
    rename(
        ascvd_date = date,
        ascvd_array_index = array_index,
        ascvd_source = dia_proc
    ) %>%
    arrange(participant_id)  # Final sorting by participant_id

```

# Join ascvd_col to the main data and fill in the missing information

Censoring date set to 2022-05-31:

```{r}

dataset_ascvd <- dataset %>% 
  select(-starts_with("diagnoses"), -starts_with("date_of_first"), -starts_with("operative_procedures"), date_of_death = date_of_death_instance_0, -date_of_death_instance_1) %>%
  left_join(ascvd_col) %>% 
  mutate(date_lost_to_follow_up = as_date(date_lost_to_follow_up),
         date_of_death =as_date(date_of_death),
    ascvd = case_when(
    is.na(ascvd) ~ 0,
    ascvd_date > as_date("2022-05-31") ~ 0,
    ascvd == 1 & date_lost_to_follow_up < ascvd_date ~ 0,
    TRUE ~ ascvd),
    ascvd_date = case_when(
      ascvd == 0 & !is.na(date_of_death) ~ date_of_death,
      ascvd == 0 & is.na(date_of_death) ~ as_date("2022-05-31"),
      is.na(ascvd_date) & ascvd == 1 ~ NA,
      TRUE ~ ascvd_date
    ),
    ascvd_date = case_when(
      date_lost_to_follow_up < ascvd_date ~ date_lost_to_follow_up,
      ascvd_date > as_date("2022-05-31") ~ as_date("2022-05-31"),
      TRUE ~ ascvd_date
    ) 
  ) 
```

#Upload to DNAnexus

```{r}
# write.csv(dataset_ascvd, "dataset_ascvd.csv")
# 
# system("dx upload dataset_ascvd.csv") 

```
